Gradual Soundness: Lessons from Static Python

### 2022-01-13 pj-7.1/paper.tex
pj: Meta + Menlo Park
pj: more sk + carl notes
pj: change tobias macro to allow vglue
somehow this was an error for me (texlive 2017) but not sk
pj: intro revised

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%%   ... or at least mention both, be clear the focus is Dict b/c older
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"
%%   - emphasize the PAYOFF from modeling early on, don't bury in sec 3
%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same

@@

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

@@

%% BEGIN tobias pape 2021-11-06
\makeatletter
\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
\makeatother
%% END

@@

\author[a]{Kuang-Chen Lu}
\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
\affiliation[a]{Brown University, Providence, RI, USA}
\author[a]{Ben Greenman}
\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a PLT member and a postdoc at Brown University.}
\author[b]{Carl Meyer}
\authorinfo{(\email{carljm@fb.com}) FILL}
\affiliation[b]{Meta}
\author[b]{Dino Viehland}
\authorinfo{(\email{dinoviehland@fb.com}) FILL}
\author[a]{Shriram Krishnamurthi}
\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}

@@

\begin{abstract}
%  Soundness, migratory typing.
%  Novel run-time check strategy, combines concrete and transient for ergonomics.

  \let\paragraph\abstractpart

@@

  \paragraph{Context}
  % What is the broad context of the work? What is
  % the importance of the general research area?
  Gradually-typed languages allow typed and untyped code to interoperate,
  but typically come with some kind of drawback.
  In some languages, the types are unreliable;
  in others, communication across type boundaries can be extremely expensive;
  and still others allow only limited forms of interoperability.
  The research community is actively seeking a sound, fast, and expressive
  approach to gradual typing.

  %% TODO acknowlegde the success of new language / tracing JIT approaches

@@

  \paragraph{Inquiry}
  % What problem or question does the paper
  % address? How has this problem or question been
  % addressed by others (if at all)?
  This paper describes \SP{}, a language developed by engineers at Instagram
  that has proven itself sound, fast, and reasonably expressive in production.
  \SP{}'s approach to gradual types is essentially a programmer-tunable combination of
  the \emph{concrete}\/ and \emph{transient}\/ approaches from the literature.
  Concrete types provide soundness in an efficient but difficult-to-use way.
  Transient types are sound in a shallow sense and easier to use; they help
  to bridge the gap between untyped code and typed concrete code.

@@

  \paragraph{Approach}
  % What was done that unveiled new knowledge?
  We evaluate the language in its current state
  to develope a model that capture the essence of its
  approach to gradual types.
  We draw upon personal communication, bug reports, and the \SP{}
  regression test suite to develop this model.

@@

  \paragraph{Knowledge}
  % What new facts were uncovered? If the
  % research was not results oriented, what new
  % capabilities are enabled by the work?
  Our main finding is that the \emph{gradual soundness}\/ that
  arises from a mix of concrete and transient types is an effective
  way to lower the maintenance cost of the concrete approach.
  On a more technical level, this paper contributes two artifacts:
  a model of \SP{} that passes all relevant tests from the \SP{} codebase,
  and a performance evaluation of \SP{}. The process of formalization
  has also found several errors in the implementation, inlcuding fatal errors.

@@

  \paragraph{Importance}
  % Why does this work matter?
  \SP{} is the first sound gradual language whose piece-meal application
  to a realistic codebase has consistently improved performance.
  %% Prior work on Nom and Dart 2 showed that concrete types are promising.
  Other language designers may wish to replicate its approach,
  especially those who maintain unsound type checkers for untyped languages
  such as JavaScript.

@@

Gradual typing has attracted significant interest as a solution to
the impasse between static and dynamic typing.
The premise is simple: let programmers introduce types in part of a
codebase and while leaving the rest untyped.
Run-time checks can in principle enforce the assumptions that
typed code makes about untyped components, thereby ensuring that
the types are sound no matter how the untyped code behaves.

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but are unsound [CITE].
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive types [CITE].
A few academics have studied the cost of run-time checks
in detail [CITE] and proposed implementation methods [CITE],
compiler technology [CITE],
and even weakened semantics [CITE], but these efforts have not yet
decisively closed the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types [CITE kafka nom etc], but its low performance
overhead stems from severe restrictions on untyped code.
%% In other words, concrete types impose low performance costs
%% but high migration costs.
%% ... you can have any color as long as it's black -h.ford
%% ... concrete expresses few programs, but runs them all nicely
Whether developers can accept these restrictions is unclear;
indeed, they may prefer a whole-sale migration over an awkward
gradual transition.\footnote{Successful migrations of untyped codebases
to a typed language are rare, but not unheard of.
Twitter ported its server-side code from Ruby to Scala
and Dropbox moved its core sync engine from Python mypy to Rust: see
\shorturl{http://www.}{artima.com/scalazine/articles/twitter\_on\_scala.html} and
\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}.}

@@

This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
Python along with a runtime system that uses soundness to drive optimizations.
The language design is a variation of concrete types that eases its restrictions.
Instead of asking programmers to migrate from untyped code to sound concrete types
in one leap, \SP{} offers an intermediate step via shallow types that enforce
only the top-level shape of values.
Programmers can begin with shallow claims and gradually dial up soundness.
On a suite of benchmarks and production modules, Static Python out-performs untyped
Python across the board.

\paragraph*{Contributions}

\begin{itemize}
  \item
    Evidence that gradual soundness delivers Python-level performance at a low migration cost
    and faster performance as programmers gradually incorporate sound types.
    \SP{} can be applied to a codebase with no refactoring and marginal performance overhead.
    After refactoring to sound types, \SP{} performs well on a microbenchmark
    suite and in production.
  \item
    A formal description of the \SP{} core language and an analysis of its soundness
    and optimizations.
    Soundness is gradual in the sense that some types are fully sound and others validate
    only top-level shapes.
  \item
    Discussions of the expressiveness limitations in \SP{} relative to
    idealized gradual type systems, and of the programming discipline that Instagram
    follows in their gradually-sound codebase.
\end{itemize}

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design.
In particular, the maintainers of optionally-typed languages
may find that adding transient and concrete types is a low-overhead
way to turn types into reliable claims.
Second, we want to lend focus to researchers.
Some of the restrictions that \SP{} adopts may be useful to
ground theoretical work.
Other restrictions might be lifted by future research.

\section{A Programmer's View of \SP{}}
\label{s:tour}
%% purpose = SP looks like Pyre, adds soundness to certain types

@@

@@

\section{Runtime System Highlights}
%% alt title: Runtime System Highlights
%% Non-Highlights:
%% - shadow frames = JIT feature, lite array to track Python call stack

@@

@@

@@

\section{Conclusion / Lessons / Design Principles}
\label{s:conclusion}

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%%   ... or at least mention both, be clear the focus is Dict b/c older
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"
%%   - emphasize the PAYOFF from modeling early on, don't bury in sec 3
%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
%% - "we" = all authors, NOT JUST BROWN!!!

@@

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}
\newcommand{\numbenchmark}{2}

@@

%% BEGIN tobias pape 2021-11-06
\makeatletter
\newcommand*\abstractpart[1]{\leavevmode\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
\makeatother
%% END

@@

\author[a]{Kuang-Chen Lu}
\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
\affiliation[a]{Brown University, Providence, RI, USA}
\author[a]{Ben Greenman}
\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a PLT member and a postdoc at Brown University.}
\author[b]{Carl Meyer}
\authorinfo{(\email{carljm@fb.com}) FILL}
\affiliation[b]{Meta, Menlo Park, CA, USA}
\author[b]{Dino Viehland}
\authorinfo{(\email{dinoviehland@fb.com}) FILL}
\author[a]{Shriram Krishnamurthi}
\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}

@@

\begin{abstract}
  \let\paragraph\abstractpart

@@

  \paragraph{Context}
  % What is the broad context of the work? What is
  % the importance of the general research area?
  Gradually-typed languages allow typed and untyped code to interoperate,
  but typically come with significant drawbacks.
  In some languages, the types are unreliable;
  in others, communication across type boundaries can be extremely expensive;
  and still others allow only limited forms of interoperability.
  The research community is actively seeking a sound, fast, and expressive
  approach to gradual typing.

@@

  \paragraph{Inquiry}
  % What problem or question does the paper
  % address? How has this problem or question been
  % addressed by others (if at all)?
  This paper describes \SP{}, a language developed by engineers at Instagram
  that has proven itself sound, fast, and reasonably expressive in production.
  \SP{}'s approach to gradual types is essentially a programmer-tunable combination of
  the \emph{concrete}\/ and \emph{transient}\/ approaches from the literature.
  Concrete types provide full soundness and low performance overhead, but impose
  nonlocal constraints.
  Transient types are sound in a shallow sense and easier to use; they help
  to bridge the gap between untyped code and typed concrete code.

@@

  \paragraph{Approach}
  % What was done that unveiled new knowledge?
  We evaluate the language in its current state
  and develop a model that captures the essence of its
  approach to gradual types.
  We draw upon personal communication, bug reports, and the \SP{}
  regression test suite to develop this model.

@@

  \paragraph{Knowledge}
  % What new facts were uncovered? If the
  % research was not results oriented, what new
  % capabilities are enabled by the work?
  Our main finding is that the \emph{gradual soundness}\/ that
  arises from a mix of concrete and transient types is an effective
  way to lower the maintenance cost of the concrete approach.
  We also find that method-based JIT technology can eliminate the
  costs of the transient approach.
  On a more technical level, this paper contributes two artifacts:
  a model of \SP{} that passes hundreds of tests from the \SP{} codebase,
  and a performance evaluation of \SP{}. The process of formalization
  found several errors in the implementation, including fatal errors.

@@

  \paragraph{Importance}
  % Why does this work matter?
  \SP{} is the first sound gradual language whose piece-meal application
  to a realistic codebase has consistently improved performance.
  %% Prior work on Nom and Dart 2 showed that concrete types are promising.
  Other language designers may wish to replicate its approach,
  especially those who currently maintain unsound gradual languages and are
  seeking a path to soundness.

@@

Gradual typing has attracted significant interest as a solution to
the debate between static and dynamic typing.
The premise is simple: let programmers introduce types in part of a
codebase and while leaving the rest untyped.
Run-time checks enforce the critical assumptions that typed code makes about
untyped components, thereby ensuring that the types are sound no matter how
untyped code behaves.

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but are unsound [CITE].
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive types [CITE].
A few academics have studied the cost of run-time checks
in detail [CITE] and proposed implementation methods [CITE],
compiler technology [CITE],
and even weakened semantics [CITE], but these efforts have not yet
decisively closed the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types [CITE kafka nom etc], but it imposes nonlocal
constraints on untyped code.
In particular, concrete-typed client code is incompatible with
values created by untyped code.
Rather than search for a coordinated set of edits to make concrete
types work, developers may prefer a whole-sale migration to
a typed language.\footnote{Successful migrations of untyped codebases
to a typed language are rare, but not unheard of.
Twitter ported its server-side code from Ruby to Scala~\cite{twitter-scala}
and Dropbox moved its core sync engine from Python mypy to Rust~\cite{dropbox-rust}.}

@@

This paper reports an exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
Python along with a runtime system that uses soundness to drive optimizations.
The language is staged to let programmers choose between easy migrations and
full-strength optimizations, a design that we call \emph{gradual soundness}.
To a first approximation, there are three main stages:
\begin{enumerate}
  \item
    \SP{} contains a full language of \emph{shallow types}\/ that describe idiomatic
    Python code at a coarse granularity.
    Adapting an untyped module to use shallow types requires at most a few local code
    changes.
  \item
    A small set of \emph{concrete types}\/ describe generic data structures
    that guarantee the types of their elements.
    If programmers modify their code to build concrete structures instead of Python
    ones (a potentially nonlocal change), then \SP{} can speed up reads from these
    structures.
  \item
    Additional types and flags to enable further optimizations.
    These include \emph{primitive types}\/ that enable unboxed arithmetic
    and \emph{strict module semantics}\/ that makes modules immutable.
\end{enumerate}

Over the past year, Instagram has been applying gradual soundness to its
primary web server monolith.
The results are very positive.
Although only a handful of modules use \SP{} types (hundreds among thousands),
and only a few critical modules rely on concrete types and primitive types (dozens),
global CPU usage has improved by almost 4\% according to internal profiling tools.
Type-directed optimizations outweigh the cost of enforcing soundness.
Consequently, we believe the details of the \SP{} approach are of interest to
the gradual typing community at large.

\paragraph*{Contributions}
This paper makes two contributions:

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes.
    At Instagram, the web server has become significantly more efficient
    after the application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    \numbenchmark{} public benchmarks.
    Using only shallow types, \SP{} runs on-par with Python despite the costs of
    enforcing soundness.
    With fine-tuned types, \SP{} consistently out-performs the Python baseline.

  \item
    \emph{Mechanization}.
    To validate the soundness of \SP{}, we mechanized a core language in PLT Redex
    and ran over 200 tests adapted from the \SP{} regression suite.
    The model brings across the essence of \SP{} and lends support to the claim
    that core \SP{} is indeed sound.
    Perhaps more importantly, the \emph{process} of modeling revealed 20 significant
    issues in \SP{}.
    Four of these were critical soundness bugs.

\paragraph*{Outline}

This paper begins with an informal description of \SP{} in two parts.
First, we present a user-oriented tour of the language~(\cref{s:tour}).
Second, we present the key ingredients of the runtime system, Cinder,
that supports \SP{}~(\cref{s:runtime}).
\Cref{s:model} uses a small formal model to introduce our Redex mechanization
and to convince readers that \SP{} is based on a sound core.
\Cref{s:impl} notes important aspects of \SP{} that are not covered by the model.
\Cref{s:eval} evaluates \SP{}; it reports our experience with the language in
production and on public microbenchmarks.
The paper concludes with related work~(\cref{s:related}),
future work~(\cref{s:future}), and a final summary~(\cref{s:conclusion}).

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design.
In particular, the maintainers of optionally-typed languages
may wish to focus on shallow types and JIT compilation as a
first step toward sound gradual types.
Second, we want to entice researchers.
The \SP{} type system has many restrictions that are suprising
at first glance; for example, functions are supported only by
shallow types.
Some of these restrictions might be lifted by future research.
Others might be useful to adopt in new language designs.

\section{A Tour of \SP{}}
\label{s:tour}
%% purpose = SP looks like Pyre, adds soundness to certain types

@@

%% 2022-01-12 Although the Static Python compiler can technically be
%% used apart from strict modules, and we do that in tests for faster tests, it’s
%% not sound and we don’t allow it in prod.

@@

\section{Runtime System Highlights}
\label{s:runtime}
%% alt title: Runtime System Highlights
%% Non-Highlights:
%% - shadow frames = JIT feature, lite array to track Python call stack

@@

%% 2022-01-12
%% We did have some cases where strict semantics caused a problem, but in the
%% majority of conversions it didn’t (or if it did initially, the fix was
%% trivial.) There was at least one case where I would have converted more code to
%% static but couldn’t easily do so because of strictness issues with a larger
%% framework reliant on metaclasses, and so I instead extracted the hottest code
%% paths into a new strict/static module and called them. I have a plan to go back
%% to this and improve the framework itself to be strict so more code can be
%% converted. I know of at least one other case where we did modify a framework
%% that relied on side-effecting decorators so that it could be used without the
%% side effects.

%% 2022-01-12 All SP mods are strict. Although the Static Python compiler can technically be
%% used apart from strict modules, and we do that in tests for faster tests, it’s
%% not sound and we don’t allow it in prod. The way a real module gets compiled as
%% Static Python is in fact via the strict modules import loader, which detects
%% both `import __strict__` and `import __static__` markers, treating the former
%% as strict only and the latter as both strict and static. So we don’t provide
%% any straightforward way (either in IG codebase or generally with Cinder) to
%% have a module imported as static without also being strict.

%% 2022-01-12 Looks like 3 use CheckedDict and 6 use CheckedList.
%% 2022-01-12

@@

%% 2022-01-12
%% I don’t think it was a super common change, but there were some cases where we
%%     changed a @classmethod that didn’t actually use its cls argument to a
%%     @staticmethod. It looks like we have 24 static modules using a
%%     @staticmethod somewhere in the module, but I think in the majority of those
%%     there just happened to be some pre-existing staticmethods (they aren’t
%%     uncommon in our codebase); I think a minority were intentional
%%     optimizations on our part. And the pre-existing ones may or may not be
%%     meaningfully improving performance. So I’m not sure how valuable it is to
%%     report this metric.

@@

\section{The Lessons}
\label{s:conclusion}

In conclusion, we summarize the
